---
title: "Graph Filters and Wavelets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Graph Filters and Wavelets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(rgsp)
set.seed(42)
```

## Overview

Graph filtering is the core operation in graph signal processing. rgsp provides:

- **Spectral kernels**: Functions that define filter response in the frequency domain
- **Chebyshev approximation**: Fast polynomial filtering without eigendecomposition
- **Filter banks**: Multiple filters applied simultaneously
- **Wavelets**: Multi-scale analysis via SGWT

## Spectral Filtering Basics

A graph filter modifies a signal based on its spectral content. Given a signal $x$ and filter kernel $h(\lambda)$:

$$y = h(L)x = U \cdot \text{diag}(h(\lambda_1), ..., h(\lambda_n)) \cdot U^T x$$

where $L = U \Lambda U^T$ is the Laplacian eigendecomposition.

### Direct vs Chebyshev Filtering

**Direct filtering** requires full eigendecomposition - expensive for large graphs:

```{r direct, eval=FALSE}
# Compute eigenpairs (expensive for large n)
eig <- graph_eigenpairs(g)
# Apply filter in spectral domain
y <- eig$vectors %*% diag(h(eig$values)) %*% t(eig$vectors) %*% x
```
**Chebyshev filtering** approximates $h(L)$ with polynomials - much faster:

```{r chebyshev}
g <- graph_ring(100)
x <- rnorm(g$n)

# Chebyshev approximation (default)
y <- filter_signal(g, x, kernel_heat(1), K = 30)
```

The parameter `K` controls accuracy (higher = more accurate but slower).

## Available Kernels

### Heat Kernel

Smoothing filter: $h(\lambda) = e^{-\tau \lambda}$

```{r kernel-heat}
g <- graph_grid2d(10, 10)
x <- numeric(g$n)
x[45] <- 1  # Impulse at center

# Different diffusion times
y1 <- filter_signal(g, x, kernel_heat(0.5), K = 30)
y2 <- filter_signal(g, x, kernel_heat(2.0), K = 30)
y3 <- filter_signal(g, x, kernel_heat(5.0), K = 30)

par(mfrow = c(1, 3))
image(matrix(y1, 10, 10), main = "tau = 0.5")
image(matrix(y2, 10, 10), main = "tau = 2.0")
image(matrix(y3, 10, 10), main = "tau = 5.0")
```

### Mexican Hat Wavelet

Band-pass filter centered at scale $\sigma$:

```{r kernel-mexican}
# Mexican hat at different scales
k1 <- kernel_mexican_hat(0.5)
k2 <- kernel_mexican_hat(1.0)
k3 <- kernel_mexican_hat(2.0)

# Evaluate on spectrum
lambda <- seq(0, 4, length.out = 100)
plot(lambda, k1(lambda), type = "l", col = 1,
     ylab = "Response", main = "Mexican Hat Kernels")
lines(lambda, k2(lambda), col = 2)
lines(lambda, k3(lambda), col = 3)
legend("topright", c("sigma=0.5", "sigma=1", "sigma=2"), col = 1:3, lty = 1)
```

### Band-pass Filters

```{r kernel-bandpass}
# Rectangle (sharp cutoff)
k_rect <- kernel_rectangle(1, 2)

# Half-cosine (smooth transition)
k_cos <- kernel_half_cosine(1, 2)

# Meyer (infinitely smooth)
k_meyer <- kernel_meyer(1, 2)

lambda <- seq(0, 4, length.out = 200)
plot(lambda, k_rect(lambda), type = "l", col = 1, ylim = c(0, 1.1),
     ylab = "Response", main = "Band-pass Filters [1, 2]")
lines(lambda, k_cos(lambda), col = 2)
lines(lambda, k_meyer(lambda), col = 3)
legend("topright", c("Rectangle", "Half-cosine", "Meyer"), col = 1:3, lty = 1)
```

### Other Kernels

```{r other-kernels, eval=FALSE}
kernel_exponential(tau)    # Exponential: tau * lambda * exp(-tau * lambda)
kernel_gabor(center, sigma) # Gabor: localized in frequency
kernel_wave(speed)          # Wave equation kernel
kernel_tight_band(a, b)     # Tight frame band-pass
```

## Filter Banks

Apply multiple filters simultaneously:

```{r filter-bank}
g <- graph_ring(64)
x <- rnorm(g$n)

# Define filter bank
bank <- list(
  lowpass = kernel_rectangle(0, 1),
  midpass = kernel_half_cosine(1, 2.5),
  highpass = kernel_rectangle(2.5, 4)
)

# Apply all filters
results <- filter_bank_apply(g, x, bank, K = 40)
str(results)
```

### Wavelet Banks

Pre-defined wavelet filter banks:

```{r wavelet-banks}
g <- graph_sensor(50, seed = 42)

# Mexican hat wavelets (4 scales)
bank_mh <- mexican_hat_wavelet_bank(g, n_scales = 4)

# Meyer wavelets
bank_meyer <- meyer_wavelet_bank(g, n_scales = 4)

# Heat diffusion wavelets (requires explicit scales)
bank_heat <- wavelet_heat_bank(scales = c(1, 2, 4, 8))

cat("Mexican hat bank:", length(bank_mh), "filters\n")
cat("Meyer bank:", length(bank_meyer), "filters\n")
```

### Tight Frame Banks

Tight frames ensure perfect reconstruction with $A = B$ (frame bounds equal):

```{r tight-frames}
g <- graph_ring(32)

# Different tight frame constructions
tf_regular <- tight_frame_regular(g)
tf_held <- tight_frame_held(g)
tf_simoncelli <- tight_frame_simoncelli(g)
tf_papadakis <- tight_frame_papadakis(g)

cat("Regular:", length(tf_regular), "filters\n")
cat("Held:", length(tf_held), "filters\n")
cat("Simoncelli:", length(tf_simoncelli), "filters\n")
cat("Papadakis:", length(tf_papadakis), "filters\n")
```

## Spectral Graph Wavelet Transform (SGWT)

The SGWT provides multi-scale analysis:

```{r sgwt-basic}
g <- graph_sensor(100, seed = 42)
set.seed(42)
x <- rnorm(g$n)

# Forward transform with 4 scales
W <- sgwt(g, x, scales = 4)
print(W)

# Access coefficients by scale
# Use "lowpass" for lowpass band, or numeric scale values for wavelet bands
lowpass_coeffs <- sgwt_coeffs_at_scale(W, "lowpass")
# Access a wavelet scale (use W$scales to see available scale values)
cat("Available scales:", W$scales, "\n")
scale1_coeffs <- sgwt_coeffs_at_scale(W, W$scales[1])

# Inverse transform
x_rec <- isgwt(W)
max(abs(x - x_rec))  # Perfect reconstruction
```

### Analyzing Wavelet Coefficients

```{r sgwt-analysis}
# Energy by scale
energy <- band_energy_df(W)
print(energy)

# Tidy format for analysis
tidy_df <- sgwt_coeffs_tidy(W)
head(tidy_df)
```

### Selective Reconstruction

Reconstruct using only certain scales:

```{r selective-recon}
# Reconstruct from wavelets only (exclude lowpass)
x_wavelets <- isgwt(W, bands = "wavelets")

# Reconstruct from lowpass only
x_lowpass <- isgwt(W, bands = "lowpass")

# Exclude lowpass using string identifier
x_no_lowpass <- isgwt(W, exclude_bands = "lowpass")
```

### Node-level Analysis

```{r node-analysis}
# Wavelet atom at specific node (takes graph, not W)
# Use one of the scales from W$scales
atom <- sgwt_atom(g, node = 50, scale = W$scales[1])

# Per-node contribution across scales (all nodes)
contrib <- sgwt_node_contribution(W)
# Show contribution for node 50
cat("Contribution for node 50:\n")
print(contrib[50, ])
```

## Heat Propagation

Simulate diffusion on the graph:

```{r heat-propagate}
g <- graph_grid2d(15, 15)
x0 <- numeric(g$n)
x0[112] <- 1  # Initial heat at center

# Propagate at multiple times (returns a list)
times <- c(0.5, 1, 2, 5)
heat <- heat_propagate(g, x0, t = times, K = 50)

par(mfrow = c(2, 2))
for (i in seq_along(times)) {
  image(matrix(heat[[i]], 15, 15), main = paste("t =", times[i]))
}
```

## Frame Analysis

Compute frame bounds to verify filter bank quality:

```{r frame-bounds}
g <- graph_ring(32)
kernels <- mexican_hat_wavelet_bank(g, n_scales = 4)

# Compute frame bounds
frame <- compute_frame(g, kernels, method = "exact")
cat("Lower bound A:", round(frame$A, 4), "\n")
cat("Upper bound B:", round(frame$B, 4), "\n")
cat("Condition number B/A:", round(frame$B / frame$A, 4), "\n")
```

A tight frame has $A \approx B$.

## Spectrogram

Vertex-frequency analysis:

```{r spectrogram}
g <- graph_ring(50)
x <- sin(seq(0, 4*pi, length.out = g$n)) + 0.5 * cos(seq(0, 8*pi, length.out = g$n))

# Compute spectrogram
spec <- compute_spectrogram(g, x, scales = c(0.5, 1, 2, 4, 8))
dim(spec)  # nodes x scales
```

## Visualization

```{r viz, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)

# Filter response plot
kernels <- list(
  Heat_1 = kernel_heat(1),
  Heat_5 = kernel_heat(5),
  MexicanHat = kernel_mexican_hat(1)
)
plot_filter_response(kernels, lmax = 4)
```

```{r viz-wavelet, eval=requireNamespace("ggplot2", quietly = TRUE)}
# Wavelet coefficient heatmap
g <- graph_ring(30)
W <- sgwt(g, sin(seq(0, 2*pi, length.out = 30)), scales = 4)
plot_wavelet_coeffs(W)
```

```{r viz-energy, eval=requireNamespace("ggplot2", quietly = TRUE)}
# Energy distribution
plot_band_energy(W)
```

## Performance Considerations

### Chebyshev Order K

- **K = 20-30**: Fast, good for smooth kernels
- **K = 50**: Good balance for most applications
- **K = 100+**: High accuracy, slower

```{r K-comparison}
g <- graph_ring(100)
x <- rnorm(g$n)
kernel <- kernel_heat(1)

# Compare different K values
y30 <- filter_signal(g, x, kernel, K = 30)
y50 <- filter_signal(g, x, kernel, K = 50)
y100 <- filter_signal(g, x, kernel, K = 100)

# Difference from high-accuracy result
cat("K=30 vs K=100 max diff:", max(abs(y30 - y100)), "\n")
cat("K=50 vs K=100 max diff:", max(abs(y50 - y100)), "\n")
```

### Caching

Chebyshev coefficients are cached automatically:

```{r caching}
g <- graph_ring(100)
x <- rnorm(g$n)
kernel <- kernel_heat(1)

# First call computes coefficients
system.time(filter_signal(g, x, kernel, K = 50))

# Second call uses cache
system.time(filter_signal(g, x, kernel, K = 50))

# Clear cache if needed
cheby_cache_clear()
```

### Lanczos Alternative

For ill-conditioned problems, try Lanczos filtering:

```{r lanczos}
g <- graph_ring(100)
x <- rnorm(g$n)

# Lanczos-based filtering
y_lanczos <- filter_signal_lanczos(g, x, kernel_heat(1), K = 30)
```

## Summary

| Task | Function | Key Parameters |
|------|----------|----------------|
| Single filter | `filter_signal()` | `kernel`, `K` |
| Multiple filters | `filter_bank_apply()` | `bank`, `K` |
| Wavelets | `sgwt()`, `isgwt()` | `scales` |
| Heat diffusion | `heat_propagate()` | `t`, `K` |
| Frame analysis | `compute_frame()` | `method` |
| Spectrogram | `compute_spectrogram()` | `scales` |
