---
title: Graph Filters and Wavelets
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Graph Filters and Wavelets} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)

library(rgsp)

set.seed(42)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Graph filtering is what you use when a node signal is easier to understand in terms of smooth versus local variation than in terms of raw adjacency entries. In `rgsp`, you define a spectral kernel, apply it with `filter_signal()`, and then inspect how the result changes across scales or bands.

This vignette uses one grid graph throughout so the effect of each filter remains visually interpretable.

## What are we filtering?

```{r build-grid}
g <- graph_grid2d(16, 16)
coords <- g$coords

impulse <- rep(0, g$n)
impulse[121] <- 1

pattern <- sin(coords[, 1] / 2) + 0.4 * cos(coords[, 2]) + 0.2 * (-1)^coords[, 1]
data.frame(nodes = g$n, edges = Matrix::nnzero(g$adjacency) / 2)
```

The `impulse` signal shows how a filter spreads information from one node to its neighbors. The `pattern` signal mixes smooth structure with faster oscillation.

## What does a heat filter do?

```{r heat-filter}
heat_small <- filter_signal(g, impulse, kernel_heat(0.2), K = 30)
heat_large <- filter_signal(g, impulse, kernel_heat(1.2), K = 30)

data.frame(
  tau = c(0.2, 1.2),
  max_value = c(max(heat_small), max(heat_large)),
  energy = c(sum(heat_small^2), sum(heat_large^2))
)
```

```{r check-heat-filter, include = FALSE}
stopifnot(
  all(is.finite(heat_small)),
  all(is.finite(heat_large)),
  max(heat_large) < max(heat_small)
)
```

A small diffusion time keeps the impulse localized. A larger diffusion time spreads the same mass more broadly over the graph.

```{r plot-heat-small, eval = has_ggplot2}
plot_signal(g, heat_small, title = "Heat filter with tau = 0.2")
```

```{r plot-heat-large, eval = has_ggplot2}
plot_signal(g, heat_large, title = "Heat filter with tau = 1.2")
```

## How do different kernels behave in frequency?

```{r compare-kernels}
kernels <- list(
  Heat = kernel_heat(0.5),
  MexicanHat = kernel_mexican_hat(1),
  BandPass = kernel_half_cosine(0.8, 2.2)
)

data.frame(
  kernel = names(kernels),
  response_at_zero = vapply(kernels, function(k) k(0), numeric(1)),
  response_at_lmax = vapply(kernels, function(k) k(lambda_max(g)), numeric(1))
)
```

```{r plot-filter-response, eval = has_ggplot2}
plot_filter_response(
  kernels,
  lmax = lambda_max(g),
  title = "Spectral responses of three common kernels"
)
```

The heat kernel is low-pass, the Mexican hat emphasizes an interior band, and the half-cosine kernel makes the band limits explicit.

## How do filter banks split a signal?

```{r filter-bank}
bank <- list(
  lowpass = kernel_rectangle(0, 0.8),
  midpass = kernel_half_cosine(0.8, 1.8),
  highpass = kernel_rectangle(1.8, lambda_max(g))
)

parts <- filter_bank_apply(g, pattern, bank, K = 40)
band_summary <- data.frame(
  band = names(parts),
  energy = vapply(parts, function(z) mean(z^2), numeric(1))
)
band_summary
```

```{r check-filter-bank, include = FALSE}
stopifnot(
  all(vapply(parts, function(z) all(is.finite(z)), logical(1))),
  all(band_summary$energy > 0)
)
```

This is the practical use of a filter bank: you get separate signals for broad trend, intermediate structure, and fine detail.

```{r plot-lowpass, eval = has_ggplot2}
plot_signal(g, parts$lowpass, title = "Low-pass component")
```

```{r plot-highpass, eval = has_ggplot2}
plot_signal(g, parts$highpass, title = "High-pass component")
```

## How do graph wavelets add scale?

```{r wavelet-analysis}
W <- sgwt(g, pattern, wavelet = "heat", scales = c(0.5, 1, 2, 4), K = 30)
energy_df <- band_energy_df(W)
energy_df[, c("band", "pct")]
```

```{r check-wavelet-analysis, include = FALSE}
pattern_rec <- isgwt(W, method = "iterative", tol = 1e-8, maxiter = 200)
stopifnot(
  all(is.finite(W$coeffs)),
  max(abs(pattern_rec - pattern)) < 1e-4
)
```

Wavelets go beyond a three-band filter bank because they let you see how much structure lives at each scale, node by node.

```{r plot-band-energy, eval = has_ggplot2}
plot_band_energy(W, title = "Wavelet energy by band")
```

```{r plot-wavelet-heatmap, eval = has_ggplot2}
plot_wavelet_coeffs(W, title = "Wavelet coefficients on the grid signal")
```

## When should you care about `strategy`?

For most users, `filter_signal()` with `strategy = "auto"` is the right default. It chooses a fast exact path on structured graphs such as rings, paths, and 2D grids, and otherwise falls back to polynomial filtering.

```{r strategy-check}
y_auto <- filter_signal(g, pattern, kernel_heat(0.4), strategy = "auto")
y_exact <- filter_signal_exact(g, pattern, kernel_heat(0.4))

data.frame(
  max_abs_difference = signif(max(abs(y_auto - y_exact)), 3)
)
```

```{r check-strategy, include = FALSE}
stopifnot(max(abs(y_auto - y_exact)) < 1e-8)
```

That means you can usually start with the default and only think about the approximation method when you are profiling or comparing algorithms.

## What should you use next?

- Use `filter_signal()` for one kernel and one task such as smoothing or band isolation.
- Use `filter_bank_apply()` when you want multiple components at once.
- Use `sgwt()` and `isgwt()` when scale-localized analysis matters more than a single filtered output.
- Use `plot_filter_response()` and `plot_wavelet_coeffs()` when you need to explain a kernel choice to yourself or someone else.
