---
title: Getting Started with rgsp
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Getting Started with rgsp} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)

library(rgsp)

set.seed(42)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

If you have one measurement per node and a graph that tells you which nodes are related, `rgsp` gives you a practical workflow for smoothing noise, inspecting graph-frequency content, and separating coarse from fine structure. In this vignette, you will build a graph, attach a noisy signal, denoise it with a heat filter, and then look at the same signal in the spectral and wavelet domains.

The main object is a `gsp_graph`: it stores the adjacency structure, coordinates when available, and cached spectral quantities. A graph signal is just a numeric vector with one value per node.

## Hero example: heat diffusion on a road network

Before getting into the mechanics, here is the kind of picture `rgsp` is built to produce. We drop a unit impulse at one node of the Minnesota road network (2,642 intersections, 3,304 roads) and watch the heat kernel spread it outward for four diffusion times. The same filter call that produces these pictures is the one you will use on your own graph a few sections down.

```{r hero-minnesota, eval = has_ggplot2, fig.width = 8, fig.height = 6.5}
library(ggplot2)

mn <- graph_minnesota()
source_node <- which.min(rowSums(sweep(mn$coords, 2, c(-93.25, 44.98))^2))  # near Minneapolis
delta <- numeric(mn$n); delta[source_node] <- 1
taus <- c(5, 25, 100, 400)

diffused <- lapply(taus, function(tau) {
  filter_signal(mn, delta, kernel_heat(tau), K = 60)
})

tri <- Matrix::summary(Matrix::tril(mn$adjacency))
edges_df <- data.frame(
  x = mn$coords[tri$i, 1], y = mn$coords[tri$i, 2],
  xend = mn$coords[tri$j, 1], yend = mn$coords[tri$j, 2]
)

nodes_df <- do.call(rbind, lapply(seq_along(taus), function(k) {
  data.frame(
    x = mn$coords[, 1], y = mn$coords[, 2],
    heat = log1p(pmax(diffused[[k]], 0) * 1e4),
    panel = factor(sprintf("tau = %g", taus[k]),
                   levels = sprintf("tau = %g", taus))
  )
}))

ggplot(nodes_df, aes(x, y)) +
  geom_segment(data = edges_df, aes(xend = xend, yend = yend),
               color = "grey85", linewidth = 0.2) +
  geom_point(aes(color = heat, alpha = heat), size = 0.45) +
  scale_color_gradientn(
    colors = c("grey80", "grey70", "#fdae61", "#d73027", "#7f0000"),
    guide = "none"
  ) +
  scale_alpha_continuous(range = c(0.25, 1), guide = "none") +
  facet_wrap(~ panel, nrow = 2) +
  coord_fixed() +
  labs(title = "Heat kernel applied to an impulse on the Minnesota road network",
       subtitle = "Same graph, same source node, four diffusion times") +
  theme_minimal(base_size = 11) +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold"),
        strip.text = element_text(face = "bold"))
```

One function call (`filter_signal`) implements the filter; the only thing that changes between panels is the scale parameter of `kernel_heat()`. Larger scales produce a more delocalized result, which is exactly what "low-pass on a graph" means when the domain is a road network instead of a regular grid.

## What do the inputs look like?

```{r build-graph}
g <- graph_sensor(80, k = 6, seed = 42)
g <- graph_set_coords(g, g$coords)

data.frame(
  nodes = g$n,
  edges = Matrix::nnzero(g$adjacency) / 2,
  has_coords = !is.null(g$coords)
)
```

Here the graph comes from a synthetic sensor network. The node coordinates let you plot the graph and any signal defined on it.

## How do you denoise a signal on a graph?

```{r make-signal}
eig <- graph_eigenpairs(g, k = 6)
x_true <- eig$vectors[, 2]
x_noisy <- x_true + rnorm(g$n, sd = 0.25)
x_smooth <- filter_signal(g, x_noisy, kernel_heat(0.2), K = 30)

data.frame(
  signal = c("noisy", "smoothed"),
  mse = c(
    mean((x_noisy - x_true)^2),
    mean((x_smooth - x_true)^2)
  )
)
```

The heat kernel suppresses high graph frequencies, so the smoothed signal keeps the broad structure of the latent pattern while discarding much of the local noise.

```{r check-denoise, include = FALSE}
stopifnot(
  all(is.finite(x_smooth)),
  mean((x_smooth - x_true)^2) < mean((x_noisy - x_true)^2)
)
```

```{r plot-denoise, eval = has_ggplot2}
plot_signal(g, x_noisy, title = "Noisy signal")
```

```{r plot-smoothed, eval = has_ggplot2}
plot_signal(g, x_smooth, title = "Heat-smoothed signal")
```

## What does the graph Fourier transform tell you?

The graph Fourier transform maps a node-domain signal onto Laplacian eigenvectors. Low-index coefficients correspond to smoother variation on the graph.

```{r graph-fourier}
x_hat <- gft(g, x_smooth)
energy_fraction <- sum(Mod(x_hat[1:10])^2) / sum(Mod(x_hat)^2)

data.frame(
  first_10_coeff_energy = round(energy_fraction, 3),
  reconstruction_error = signif(max(abs(igft(g, x_hat) - x_smooth)), 3)
)
```

```{r check-gft, include = FALSE}
stopifnot(
  all(is.finite(x_hat)),
  max(abs(igft(g, x_hat) - x_smooth)) < 1e-8
)
```

This gives you a quick diagnostic: if most of the energy sits in the low frequencies, the signal is graph-smooth.

## How do wavelets separate coarse and fine structure?

```{r wavelet-transform}
W <- sgwt(g, x_smooth, scales = c(0.5, 1, 2, 4), K = 30)
energy_df <- band_energy_df(W)
energy_df[, c("band", "pct")]
```

```{r check-wavelets, include = FALSE}
x_rec <- isgwt(W, method = "iterative", tol = 1e-8, maxiter = 200)
stopifnot(
  all(is.finite(W$coeffs)),
  max(abs(x_rec - x_smooth)) < 1e-4
)
```

The lowpass band summarizes the broad background trend, while the wavelet bands capture increasingly local structure.

```{r plot-band-energy, eval = has_ggplot2}
plot_band_energy(W, title = "Where does the wavelet energy live?")
```

```{r plot-wavelet-coeffs, eval = has_ggplot2}
plot_wavelet_coeffs(W, title = "Wavelet coefficients across nodes and scales")
```

## Which graph constructors should you reach for?

For most workflows, you only need one of a few constructors:

- `graph_sensor()` for point clouds and spatial networks
- `graph_ring()` and `graph_path()` for 1D toy problems
- `graph_grid2d()` for image-like lattices
- `graph_sbm()` and `graph_random_geometric()` for controlled random examples

Once you have a graph, the common next functions are `filter_signal()`, `gft()`, `sgwt()`, and `tikhonov_smooth()`.

## Where do you go next?

- Use `vignette("filters")` when you want to compare kernels, filter banks, and wavelets in more detail.
- Use `vignette("optimization")` when your main task is denoising, interpolation, or graph-based learning.
- Use `plot_graph()`, `plot_signal()`, and `plot_filter_response()` when you need quick diagnostics rather than raw object inspection.
