---
title: Getting Started with rgsp
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Getting Started with rgsp} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)

library(rgsp)

set.seed(42)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

If you have one measurement per node and a graph that tells you which nodes are related, `rgsp` gives you a practical workflow for smoothing noise, inspecting graph-frequency content, and separating coarse from fine structure. In this vignette, you will build a graph, attach a noisy signal, denoise it with a heat filter, and then look at the same signal in the spectral and wavelet domains.

The main object is a `gsp_graph`: it stores the adjacency structure, coordinates when available, and cached spectral quantities. A graph signal is just a numeric vector with one value per node.

## What do the inputs look like?

```{r build-graph}
g <- graph_sensor(80, k = 6, seed = 42)
g <- graph_set_coords(g, g$coords)

data.frame(
  nodes = g$n,
  edges = Matrix::nnzero(g$adjacency) / 2,
  has_coords = !is.null(g$coords)
)
```

Here the graph comes from a synthetic sensor network. The node coordinates let you plot the graph and any signal defined on it.

## How do you denoise a signal on a graph?

```{r make-signal}
eig <- graph_eigenpairs(g, k = 6)
x_true <- eig$vectors[, 2]
x_noisy <- x_true + rnorm(g$n, sd = 0.25)
x_smooth <- filter_signal(g, x_noisy, kernel_heat(0.2), K = 30)

data.frame(
  signal = c("noisy", "smoothed"),
  mse = c(
    mean((x_noisy - x_true)^2),
    mean((x_smooth - x_true)^2)
  )
)
```

The heat kernel suppresses high graph frequencies, so the smoothed signal keeps the broad structure of the latent pattern while discarding much of the local noise.

```{r check-denoise, include = FALSE}
stopifnot(
  all(is.finite(x_smooth)),
  mean((x_smooth - x_true)^2) < mean((x_noisy - x_true)^2)
)
```

```{r plot-denoise, eval = has_ggplot2}
plot_signal(g, x_noisy, title = "Noisy signal")
```

```{r plot-smoothed, eval = has_ggplot2}
plot_signal(g, x_smooth, title = "Heat-smoothed signal")
```

## What does the graph Fourier transform tell you?

The graph Fourier transform maps a node-domain signal onto Laplacian eigenvectors. Low-index coefficients correspond to smoother variation on the graph.

```{r graph-fourier}
x_hat <- gft(g, x_smooth)
energy_fraction <- sum(Mod(x_hat[1:10])^2) / sum(Mod(x_hat)^2)

data.frame(
  first_10_coeff_energy = round(energy_fraction, 3),
  reconstruction_error = signif(max(abs(igft(g, x_hat) - x_smooth)), 3)
)
```

```{r check-gft, include = FALSE}
stopifnot(
  all(is.finite(x_hat)),
  max(abs(igft(g, x_hat) - x_smooth)) < 1e-8
)
```

This gives you a quick diagnostic: if most of the energy sits in the low frequencies, the signal is graph-smooth.

## How do wavelets separate coarse and fine structure?

```{r wavelet-transform}
W <- sgwt(g, x_smooth, scales = c(0.5, 1, 2, 4), K = 30)
energy_df <- band_energy_df(W)
energy_df[, c("band", "pct")]
```

```{r check-wavelets, include = FALSE}
x_rec <- isgwt(W, method = "iterative", tol = 1e-8, maxiter = 200)
stopifnot(
  all(is.finite(W$coeffs)),
  max(abs(x_rec - x_smooth)) < 1e-4
)
```

The lowpass band summarizes the broad background trend, while the wavelet bands capture increasingly local structure.

```{r plot-band-energy, eval = has_ggplot2}
plot_band_energy(W, title = "Where does the wavelet energy live?")
```

```{r plot-wavelet-coeffs, eval = has_ggplot2}
plot_wavelet_coeffs(W, title = "Wavelet coefficients across nodes and scales")
```

## Which graph constructors should you reach for?

For most workflows, you only need one of a few constructors:

- `graph_sensor()` for point clouds and spatial networks
- `graph_ring()` and `graph_path()` for 1D toy problems
- `graph_grid2d()` for image-like lattices
- `graph_sbm()` and `graph_random_geometric()` for controlled random examples

Once you have a graph, the common next functions are `filter_signal()`, `gft()`, `sgwt()`, and `tikhonov_smooth()`.

## Where do you go next?

- Use `vignette("filters")` when you want to compare kernels, filter banks, and wavelets in more detail.
- Use `vignette("optimization")` when your main task is denoising, interpolation, or graph-based learning.
- Use `plot_graph()`, `plot_signal()`, and `plot_filter_response()` when you need quick diagnostics rather than raw object inspection.
