---
title: "Getting Started with rgsp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rgsp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**rgsp** is an R package for Graph Signal Processing (GSP), providing tools to analyze and process signals defined on graphs. It is a port of the Python [PyGSP](https://pygsp.readthedocs.io/) library with C++ acceleration via RcppArmadillo.

Key features:

- **Graph construction**: Ring, path, grid, sensor networks, random graphs
- **Spectral analysis**: Graph Fourier Transform (GFT), eigendecomposition
- **Filtering**: Chebyshev polynomial approximation, various kernels
- **Wavelets**: Spectral Graph Wavelet Transform (SGWT)
- **Optimization**: Tikhonov regularization, total variation
- **Visualization**: ggplot2-based plotting

```{r setup}
library(rgsp)
```

## Creating Graphs

### Built-in Graph Types

rgsp provides many graph constructors:
```{r graphs}
# Simple graphs
g_ring <- graph_ring(20)
g_path <- graph_path(20)
g_grid <- graph_grid2d(5, 4)

# Random graphs
g_er <- graph_erdos_renyi(30, p = 0.2, seed = 42)
g_sensor <- graph_sensor(50, seed = 42)

# Structured graphs
g_complete <- graph_complete(10)
# SBM requires a probability matrix P (k x k for k communities)
P <- matrix(c(0.5, 0.05, 0.05,
              0.05, 0.5, 0.05,
              0.05, 0.05, 0.5), nrow = 3)
g_sbm <- graph_sbm(c(10, 10, 10), P = P, seed = 42)
```

### Graph Properties

Each graph object contains:

```{r properties}
g <- graph_ring(10)
g$n          # Number of nodes
g$W[1:5, 1:5]  # Adjacency matrix (sparse)

# Compute Laplacian
L <- graph_laplacian(g)

# Get node degrees
deg <- graph_degree(g)
print(deg)
```

## Signals on Graphs

A graph signal is simply a vector with one value per node:

```{r signals}
g <- graph_ring(50)

# Create a smooth signal
x_smooth <- sin(seq(0, 2*pi, length.out = g$n))

# Create a noisy signal
set.seed(42)
x_noisy <- x_smooth + rnorm(g$n, sd = 0.3)
```

## Graph Fourier Transform

The GFT projects signals onto the graph's eigenvector basis:

```{r gft}
g <- graph_ring(32)
x <- sin(seq(0, 4*pi, length.out = g$n))

# Forward transform
x_hat <- gft(g, x)

# Inverse transform (perfect reconstruction)
x_rec <- igft(g, x_hat)
max(abs(x - x_rec))  # Should be ~0
```

## Filtering Signals

### Heat Kernel Filter

The heat kernel smooths signals on the graph:

```{r filter-heat}
g <- graph_ring(64)
set.seed(42)
x <- rnorm(g$n)  # Random signal

# Apply heat filter (smoothing)
y <- filter_signal(g, x, kernel_heat(t = 1.0), K = 30)

# Compare
plot(x, type = "l", col = "gray", main = "Heat Filtering")
lines(y, col = "blue", lwd = 2)
legend("topright", c("Original", "Filtered"), col = c("gray", "blue"), lwd = c(1, 2))
```

### Available Kernels

```{r kernels, eval=FALSE}
kernel_heat(tau)           # Heat diffusion: exp(-tau * lambda)
kernel_exponential(tau)    # Exponential decay
kernel_mexican_hat(sigma)  # Mexican hat wavelet
kernel_rectangle(a, b)     # Band-pass [a, b]
kernel_half_cosine(a, b)   # Smooth band-pass
kernel_meyer(a, b)         # Meyer wavelet kernel
```

## Spectral Graph Wavelets

The SGWT decomposes signals across multiple scales:

```{r sgwt}
g <- graph_sensor(100, seed = 42)
set.seed(42)
x <- rnorm(g$n)

# Compute wavelet coefficients
W <- sgwt(g, x, scales = 4)
print(W)

# Reconstruct signal
x_rec <- isgwt(W)
max(abs(x - x_rec))  # Perfect reconstruction
```

### Wavelet Energy Analysis

```{r wavelet-energy}
# Energy distribution across scales
energy_df <- band_energy_df(W)
print(energy_df)
```

## Visualization

rgsp provides ggplot2-based plotting (requires ggplot2 installed):

```{r plotting, eval=requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)

# Create a graph with coordinates
g <- graph_sensor(50, seed = 42)
g <- graph_set_coords(g, "spring", seed = 123)

# Plot graph structure
plot_graph(g)
```

```{r plot-signal, eval=requireNamespace("ggplot2", quietly = TRUE)}
# Plot signal on graph
signal <- sin(g$coords[,1] * 3)  # Signal based on x-coordinate
plot_signal(g, signal)
```

```{r plot-filter, eval=requireNamespace("ggplot2", quietly = TRUE)}
# Plot filter responses
kernels <- list(
  Heat = kernel_heat(1),
  Mexican = kernel_mexican_hat(1)
)
plot_filter_response(kernels, lmax = 4)
```

## Optimization

### Tikhonov Smoothing

```{r tikhonov}
g <- graph_ring(50)
set.seed(42)
x_true <- sin(seq(0, 2*pi, length.out = g$n))
x_noisy <- x_true + rnorm(g$n, sd = 0.3)

# Smooth with Tikhonov regularization
x_smooth <- tikhonov_smooth(g, x_noisy, tau = 2)

# Compare MSE
mse_noisy <- mean((x_noisy - x_true)^2)
mse_smooth <- mean((x_smooth - x_true)^2)
cat("MSE noisy:", round(mse_noisy, 4), "\n")
cat("MSE smooth:", round(mse_smooth, 4), "\n")
```
### Total Variation Denoising

```{r tv-denoise}
# TV denoising preserves edges better than Tikhonov
x_tv <- tv_denoise(g, x_noisy, lambda = 0.5, maxit = 100)
mse_tv <- mean((x_tv - x_true)^2)
cat("MSE TV:", round(mse_tv, 4), "\n")
```

## Performance Tips

1. **Caching**: Eigenpairs and Chebyshev coefficients are cached automatically
2. **Filter order K**: Higher K = more accurate but slower. K=30-50 usually sufficient
3. **Sparse matrices**: rgsp uses sparse matrices throughout for efficiency
4. **Clear caches** if memory is a concern:

```{r caching, eval=FALSE}
graph_cache_clear()
cheby_cache_clear()
```

## Next Steps

- See `vignette("filters")` for detailed filter and wavelet documentation
- See `vignette("optimization")` for optimization algorithms
- Check `?rgsp` for the full function reference

## Session Info

```{r session}
sessionInfo()
```
