---
title: Optimization and Learning
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html

vignette: |
  %\VignetteIndexEntry{Optimization and Learning} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
library(rgsp)
set.seed(42)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) library(ggplot2)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Overview

rgsp provides optimization algorithms for graph-based regularization:

- **Tikhonov regularization**: Smooth solutions via Laplacian penalty
- **Total variation**: Edge-preserving regularization
- **Interpolation/Inpainting**: Recover missing values
- **Proximal operators**: Building blocks for convex optimization
- **Learning**: Regression and classification on graphs

## Tikhonov Regularization

Tikhonov smoothing minimizes:

$$\min_x \|x - y\|^2 + \tau \cdot x^T L x$$

where $L$ is the graph Laplacian and $\tau$ controls smoothness.

```{r tikhonov-basic, eval = has_ggplot2}
g <- graph_ring(100)
x_true <- sin(seq(0, 2 * pi, length.out = g$n))
x_noisy <- x_true + rnorm(g$n, sd = 0.3)
x_smooth <- tikhonov_smooth(g, x_noisy, tau = 2)

tikh_df <- data.frame(
  node = rep(seq_len(g$n), 3),
  value = c(x_noisy, x_true, x_smooth),
  series = factor(rep(c("Noisy", "True", "Smoothed"), each = g$n),
                  levels = c("Noisy", "True", "Smoothed"))
)

ggplot(tikh_df, aes(node, value, color = series, linetype = series, shape = series)) +
  geom_point(data = subset(tikh_df, series == "Noisy"), size = 1.3, alpha = 0.6) +
  geom_line(data = subset(tikh_df, series != "Noisy"), linewidth = 1) +
  scale_color_manual(values = c(Noisy = "grey50", True = "black", Smoothed = "#1f78b4")) +
  scale_linetype_manual(values = c(Noisy = "blank", True = "solid", Smoothed = "solid")) +
  scale_shape_manual(values = c(Noisy = 16, True = NA, Smoothed = NA)) +
  labs(title = "Tikhonov smoothing on a ring graph", x = "node index", y = NULL,
       color = NULL, linetype = NULL, shape = NULL)
```

### Effect of Regularization Parameter

```{r tikhonov-tau, eval = has_ggplot2, fig.height = 5}
taus <- c(0.1, 1, 5, 20)
tau_df <- do.call(rbind, lapply(taus, function(tau) {
  data.frame(
    node = seq_len(g$n),
    noisy = x_noisy,
    true = x_true,
    smoothed = tikhonov_smooth(g, x_noisy, tau = tau),
    panel = factor(sprintf("tau = %g", tau), levels = sprintf("tau = %g", taus))
  )
}))

ggplot(tau_df, aes(node)) +
  geom_point(aes(y = noisy), color = "grey60", size = 0.9, alpha = 0.6) +
  geom_line(aes(y = true), color = "grey30", linewidth = 0.5) +
  geom_line(aes(y = smoothed), color = "#1f78b4", linewidth = 1) +
  facet_wrap(~ panel) +
  labs(title = "Increasing tau trades fidelity for smoothness",
       x = "node index", y = NULL)
```

## Total Variation Denoising

TV denoising minimizes:

$$\min_x \frac{1}{2}\|x - y\|^2 + \lambda \|Dx\|_1$$

where $D$ is the graph gradient operator. TV preserves edges better than Tikhonov.

```{r tv-denoise, eval = has_ggplot2}
g <- graph_path(100)
x_true <- c(rep(0, 30), rep(1, 40), rep(0, 30))
x_noisy <- x_true + rnorm(g$n, sd = 0.2)
x_tv <- tv_denoise(g, x_noisy, lambda = 1, maxit = 200)
x_tikh <- tikhonov_smooth(g, x_noisy, tau = 5)

tv_df <- data.frame(
  node = rep(seq_len(g$n), 4),
  value = c(x_noisy, x_true, x_tv, x_tikh),
  series = factor(rep(c("Noisy", "True", "TV", "Tikhonov"), each = g$n),
                  levels = c("Noisy", "True", "TV", "Tikhonov"))
)

ggplot(tv_df, aes(node, value, color = series)) +
  geom_point(data = subset(tv_df, series == "Noisy"), size = 1.2, alpha = 0.5) +
  geom_line(data = subset(tv_df, series == "True"), linewidth = 0.8) +
  geom_line(data = subset(tv_df, series == "TV"), linewidth = 1.1) +
  geom_line(data = subset(tv_df, series == "Tikhonov"), linewidth = 1.1, linetype = "dashed") +
  scale_color_manual(values = c(Noisy = "grey60", True = "black",
                                TV = "#d73027", Tikhonov = "#1f78b4")) +
  labs(title = "TV preserves the jump; Tikhonov rounds it off",
       x = "node index", y = NULL, color = NULL)
```

### TV Parameters

```{r tv-params, eval = has_ggplot2, fig.height = 5}
lambdas <- c(0.1, 0.5, 1, 2)
tv_df2 <- do.call(rbind, lapply(lambdas, function(lam) {
  data.frame(
    node = seq_len(g$n),
    noisy = x_noisy,
    true = x_true,
    denoised = tv_denoise(g, x_noisy, lambda = lam, maxit = 200),
    panel = factor(sprintf("lambda = %g", lam),
                   levels = sprintf("lambda = %g", lambdas))
  )
}))

ggplot(tv_df2, aes(node)) +
  geom_point(aes(y = noisy), color = "grey60", size = 0.9, alpha = 0.6) +
  geom_line(aes(y = true), color = "grey30", linewidth = 0.5) +
  geom_line(aes(y = denoised), color = "#d73027", linewidth = 1.1) +
  facet_wrap(~ panel) +
  labs(title = "Larger lambda collapses variation, even across the true jump",
       x = "node index", y = NULL)
```

## Inpainting (Missing Value Recovery)

### Laplacian Interpolation

Recover missing values by minimizing the Laplacian quadratic form:

```{r inpaint-laplacian}
g <- graph_grid2d(10, 10)

# Create signal with missing values
x_true <- sin(seq(0, 2*pi, length.out = g$n))
x_obs <- x_true
missing_idx <- sample(g$n, 30)  # 30% missing
x_obs[missing_idx] <- NA

# Laplacian interpolation
x_interp <- interpolate_laplacian(g, x_obs)

cat("MSE on missing values:",
    mean((x_interp[missing_idx] - x_true[missing_idx])^2), "\n")
```

### TV Inpainting

TV inpainting preserves edges in the recovered signal:

```{r inpaint-tv, eval = has_ggplot2}
x_true <- c(rep(0, 30), rep(1, 40), rep(0, 30))
x_obs <- x_true
x_obs[sample(100, 20)] <- NA  # 20% missing
g <- graph_path(100)
x_tv_inpaint <- tv_inpaint(g, x_obs, lambda = 0.5, maxit = 200)

missing_idx <- which(is.na(x_obs))
inpaint_df <- rbind(
  data.frame(node = seq_len(g$n), value = x_true,
             role = "True signal"),
  data.frame(node = missing_idx, value = x_tv_inpaint[missing_idx],
             role = "Recovered (was missing)"),
  data.frame(node = setdiff(seq_len(g$n), missing_idx),
             value = x_obs[!is.na(x_obs)],
             role = "Observed")
)
inpaint_df$role <- factor(inpaint_df$role,
  levels = c("True signal", "Observed", "Recovered (was missing)"))

ggplot(inpaint_df, aes(node, value, color = role, shape = role, size = role)) +
  geom_line(data = subset(inpaint_df, role == "True signal"), linewidth = 0.7) +
  geom_point(data = subset(inpaint_df, role != "True signal")) +
  scale_color_manual(values = c(`True signal` = "grey40",
                                Observed = "black",
                                `Recovered (was missing)` = "#d73027")) +
  scale_shape_manual(values = c(`True signal` = NA, Observed = 1,
                                `Recovered (was missing)` = 16)) +
  scale_size_manual(values = c(`True signal` = 0.7, Observed = 1.5,
                               `Recovered (was missing)` = 2.5)) +
  labs(title = "TV inpainting: 20% of nodes hidden, recovered in red",
       x = "node index", y = NULL, color = NULL, shape = NULL, size = NULL)
```

## Proximal Operators

Building blocks for convex optimization algorithms:

### L1 Proximal (Soft Thresholding)

$$\text{prox}_{\lambda \|\cdot\|_1}(x) = \text{sign}(x) \max(|x| - \lambda, 0)$$

```{r prox-l1}
x <- seq(-2, 2, length.out = 100)
y <- prox_l1(x, tau = 0.5)

plot(x, x, type = "l", col = "gray", lty = 2,
     main = "L1 Proximal (Soft Thresholding)")
lines(x, y, col = "blue", lwd = 2)
legend("topleft", c("Identity", "Soft threshold (tau=0.5)"),
       col = c("gray", "blue"), lty = c(2, 1), lwd = c(1, 2))
```

### Box Constraint Projection

```{r prox-box}
x <- rnorm(10)
y <- prox_box(x, lower = 0, upper = 1)
cbind(x = x, projected = y)
```

### Simplex Projection

Project onto the probability simplex:

```{r proj-simplex}
x <- c(-1, 0.5, 3, 2)
y <- proj_simplex(x)
cat("Input:", x, "\n")
cat("Projected:", round(y, 4), "\n")
cat("Sum:", sum(y), "\n")
```

### L2 Ball Projection

```{r proj-l2}
x <- c(3, 4)  # Norm = 5
y <- proj_l2_ball(x, r = 2)
cat("Input norm:", sqrt(sum(x^2)), "\n")
cat("Projected norm:", sqrt(sum(y^2)), "\n")
```

### 1D Total Variation Proximal

```{r prox-tv1d}
x <- c(1, 1.1, 1.2, 5, 5.1, 5.2, 2, 2.1)
y <- prox_tv1d(x, lambda = 1)
plot(x, type = "b", pch = 16, main = "1D TV Proximal")
lines(y, col = "blue", type = "b", pch = 17)
legend("topright", c("Original", "TV proximal"), col = c("black", "blue"), pch = c(16, 17))
```

## Learning on Graphs

### Tikhonov Regression

Semi-supervised regression with graph regularization:

```{r regression}
g <- graph_sensor(100, seed = 42)

# True smooth function on graph
eig <- graph_eigenpairs(g)
x_true <- eig$vectors[, 2]  # Second eigenvector (smooth)

# Observe noisy values at subset of nodes
obs_idx <- sample(g$n, 20)
y_obs <- x_true[obs_idx] + rnorm(20, sd = 0.1)

# Create observation vector (NA for unobserved)
y <- rep(NA, g$n)
y[obs_idx] <- y_obs

# Regression
x_pred <- regression_tikhonov(g, y, tau = 1)

# Evaluate
mse <- mean((x_pred - x_true)^2)
cat("MSE:", round(mse, 6), "\n")
```

### Tikhonov Classification

Semi-supervised classification:

```{r classification}
# Create SBM with probability matrix (high within-block, low between-block)
P <- matrix(c(0.3, 0.02, 0.02,
              0.02, 0.3, 0.02,
              0.02, 0.02, 0.3), nrow = 3)
g <- graph_sbm(c(30, 30, 40), P = P, seed = 42)

# True labels (0, 1, 2 for three communities)
labels_true <- c(rep(0, 30), rep(1, 30), rep(2, 40))

# Observe labels at subset
obs_idx <- c(sample(1:30, 5), sample(31:60, 5), sample(61:100, 5))
labels_obs <- rep(NA, g$n)
labels_obs[obs_idx] <- labels_true[obs_idx]

# Classification
labels_pred <- classification_tikhonov(g, labels_obs, tau = 0.1)

# Accuracy
accuracy <- mean(labels_pred == labels_true)
cat("Accuracy:", round(accuracy * 100, 1), "%\n")
```

## Graph Coarsening

### Kron Reduction

Reduce graph size while preserving spectral properties:

```{r kron}
g <- graph_grid2d(6, 6)

# Keep every other node
keep_idx <- seq(1, g$n, by = 2)
g_reduced <- kron_reduction(g, keep_idx)

cat("Original nodes:", g$n, "\n")
cat("Reduced nodes:", g_reduced$n, "\n")
```

### Multiresolution Pyramid

Create a hierarchy of coarsened graphs:

```{r multiresolution}
g <- graph_sensor(100, seed = 42)

# Create multiresolution hierarchy
hier <- graph_multiresolution(g, levels = 3)

cat("Levels:", length(hier$graphs), "\n")
for (i in seq_along(hier$graphs)) {
  cat("  Level", i, ":", hier$graphs[[i]]$n, "nodes\n")
}
```

### Pyramid Transforms

Analyze signals across resolutions:

```{r pyramid}
g <- graph_sensor(64, seed = 42)
x <- rnorm(g$n)

# Analysis: decompose into coarse + details (keep every other node)
pyr <- pyramid_analysis(g, x)

cat("Coarse signal length:", length(pyr$coarse), "values\n")
cat("Detail signal length:", length(pyr$detail), "values\n")

# Synthesis: reconstruct
x_rec <- pyramid_synthesis(pyr)
max(abs(x - x_rec))  # Perfect reconstruction
```

## Differential Operators

### Gradient and Divergence

```{r grad-div}
g <- graph_grid2d(5, 5)
x <- rnorm(g$n)

# Gradient: maps node signal to edge signal
grad_x <- grad(g, x)
cat("Signal length:", length(x), "\n")
cat("Gradient length:", length(grad_x), "\n")

# Divergence: maps edge signal back to nodes
div_grad_x <- div(g, grad_x)

# div(grad(x)) = L * x (Laplacian)
L <- graph_laplacian(g)
Lx <- as.numeric(L %*% x)
max(abs(div_grad_x - Lx))  # Should be ~0
```

### Differential Operator Matrix

```{r diff-op}
g <- graph_ring(10)
D <- compute_differential_operator(g)
dim(D)  # edges x nodes
```

## Performance Tips

### Convergence Parameters

```{r convergence}
g <- graph_ring(100)
x_noisy <- sin(seq(0, 2*pi, length.out = g$n)) + rnorm(100, sd = 0.3)

# Adjust convergence for speed vs accuracy
x_fast <- tv_denoise(g, x_noisy, lambda = 1, maxit = 50, tol = 1e-3)
x_accurate <- tv_denoise(g, x_noisy, lambda = 1, maxit = 500, tol = 1e-6)

cat("Fast iterations: 50, tol: 1e-3\n")
cat("Accurate iterations: 500, tol: 1e-6\n")
```

### Warm Starting

For multiple similar problems, use previous solution as initialization:

```{r warm-start, eval=FALSE}
# Not directly supported, but can be achieved by:
# 1. Run with fewer iterations
# 2. Use result as starting point for next problem
```

## Summary

| Task | Function | Key Parameters |
|------|----------|----------------|
| Smoothing | `tikhonov_smooth()` | `tau` |
| TV denoising | `tv_denoise()` | `lambda`, `maxit`, `tol` |
| Interpolation | `interpolate_laplacian()` | - |
| TV inpainting | `tv_inpaint()` | `lambda`, `maxit` |
| Regression | `regression_tikhonov()` | `tau` |
| Classification | `classification_tikhonov()` | `tau` |
| Coarsening | `kron_reduction()` | `keep` |
| Pyramid | `pyramid_analysis/synthesis()` | `levels` |

## References

- Shuman, D. I., et al. (2013). "The emerging field of signal processing on graphs." IEEE Signal Processing Magazine.
- Hammond, D. K., et al. (2011). "Wavelets on graphs via spectral graph theory." Applied and Computational Harmonic Analysis.
